SUPPORTING INFORMATION:\\ Headgroup Structure and Cation Binding in Phosphatidylserine Lipid Bilayers

### 2019-10-16 Manuscript/manuscriptPSsuppl.tex
Add files via upload
Modified SI and SI bibliography

Simulations of these systems were performed for 500~ns using GROMACS, version 5.0.6 \cite{abraham2015gromacs}.
A 2~fs timestep was applied during the simulations, and the LINCS algorithm was used to constrain all bonds to hydrogen atoms \cite{hess97,hess07}.
For each of the DOPS and POPS systems, two simulations were performed using different randomly assigned starting velocities.
The simulations were maintained at temperatures of 303~K and 298~K for DOPS and POPS respectively using the Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}
with a coupling constant of 1~ps. A pressure of 1~bar was maintained using the Parrinello--Rahman \cite{parrinello81} method with a coupling constant of 5~ps.
Pressure coupling was applied in a semi-isotropic manner with the $x$ and $y$ dimensions, in the plane of the bilayer, fluctuating independently of the $z$ dimension.
Standard CHARMM36 force field methods were applied for the simulation cut-offs: van der Waals interactions were truncated at 1.2~nm with the interactions
switched off between 1.0~nm and 1.2~nm; Coulombic interactions were truncated at 1.2~nm with long-range interactions treated using PME~\cite{darden93,essman95}.

Analyses on these simulations was performed on the final 100~ns of the simulations.

\noindent
{\it POPC:POPS mixtures without additional ions}
Simulations of POPC:POPS (5:1) mixtures containing a total of 132 lipids (110 POPC and 22 POPS) were constructed using the CHARMM-GUI~\cite{lee16,jo18}.
Two identical systems were constructed, apart from one system was neutralised through the addition of 22 Na$^+$ ions while the other system
contained 22 K$^+$ ions. As per the pure POPS CHARMM36 simulations, 35 water per lipid were added. All other system and simulation parameters
and settings were identical to those used in the CHARMM36 POPS simulations.

@@

Simulations of POPC:POPS (5:1) ans POPC:POPS (1:1) mixtures containing total 300 lipids (250:50 and 150:150, respectively)
with neutralizing potassium countertions 
%ran using CHARMM36 force field.
%The data are available from Refs. \citenum{POPC5POPS1noCaClCHARMM} and \citenum{POPC1POPS1noCaClCHARMM}.
%POPC:POPS (both 1:1 and 5:1) mixtures
were prepared
%with 300 lipids in total
%with the  ions
using the CHARMM-GUI \cite{lee16,jo18}. The systems were simulated using Gromacs 5 \cite{abraham2015gromacs}
and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298~K. For further details see table \ref{mixedIONsystemsMONOVALENT}. 
 \\

\noindent
{\it POPC:POPS (5:1) mixtures with additional monovalent ions.}
POPC:POPS (110:22) mixtures containing total 132 lipids with the additional
potassium or sodium ions corresponding concentrations of $\sim$450~mM and 890~mM
were generated with the CHARMM-GUI \cite{lee16,jo18}. Systems were simulated using
Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation
parameters given by the CHARMM-GUI at 298~K. For further details, simulation files
and data, see table \ref{mixedIONsystemsMONOVALENT} and
Refs. \citenum{charmm36pops+83popcT298Kwith450mMK,charmm36pops+83popcT298Kwith890mMK,charmm36pops+83popcT298Kwith440mMNa,charmm36pops+83popcT298Kwith890mMNa}. \\

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium using NBfix1 parameters.} 
POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids
%were simulated with the additional calcium concentrations.
%The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
and 0.26~M and 1~M CaCl$_2$ were prepared using the CHARMM-GUI in January 2018 \cite{lee16,jo18}.
The ion parameters generated by CHARMM-GUI at the time are labeled here as NBfix1~\cite{kim16}.
Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given
by the CHARMM-GUI at 298~K. \\

@@

\noindent
{\it POPS bilayers with Na$^+$ counterions.}
%The data in Table \ref{PSsystems} and Refs \citenum{macrogPOPS298Kcorrect,macrogPOPS298KwithK}
%\todo{T. Piggot and M. Javanainen, please write simulation details} \\
%\hl{Piggot: please write about system with Na$^+$ counterions}.
Force field parameters for POPS simulations with the MacRog OPLS-AA force field~\cite{maciejewski14,kulig15b,rog16}
were constructed manually using the provided PS parameters as a guide~\cite{rog16}. The original parameters were
not used due to an incorrect arrangement of the {\it sn}-1 and {\it sn}-2 tails (i.e. the lipid was OPPS and not POPS).
The corrected parameters are available from Ref.~\citenum{macrogPOPS298KwithK}.
The starting structure was the same as that used in the CHARMM36 POPS simulations (with Na$^+$ ions) and was converted into
the correct format using a script to adjust the atom order. Water was treated using the TIP3P model~\cite{jorgensen83}.
Simulation parameters were chosen to closely mimic those used in the original force field publications~\cite{maciejewski14,kulig15b,rog16}:
van der Waals interactions were truncated at 1.0~nm with a dispersion correction applied to the energy and pressure;
Long-range Coulombic interactions were treated with PME beyond a system-optimized cut-off.
All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
including use of the Verlet cut-off scheme~\cite{Pall13}, apart from the use of LINCS to constrain all bonds during these simulations.
Simulations with the group cut-off scheme gave slightly lower area per molecule ($\sim$~50\AA$^2$) than
the simulations with Verlet cut-off ($\sim$~52\AA$^2$).

@@

\noindent
{\it POPC:POPS (5:1) mixtures with additional potassium ions} \\
%The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsMACROG,POPCpopsMACROGwithK}
The bilayers containing a total of 120 POPC and 24 POPS molecules distributed evenly among the 
two leaflets were set up using CHARMM-GUI \cite{lee16,jo18}.
The bilayers were solvated by 5760 water molecules, corresponding to 40 water molecules for lipid.
Additionally, 24 potassium ions were added to neutralize the charge of the POPS head groups.
These bilayers were simulated using Gromacs 2016.3~\cite{abraham2015gromacs} in the presence of counterions only, as well as together with different
concentrations of CaCl$_2$ or KCl. For the former, concentrations of 100~mM, 300~mM, 1~M, and 3~M 
were considered, which corresponded to the amounts of 10/20, 31/62, 104/208, and 311/622 
of Ca$^{2+}$/Cl$^-$ ions. For membranes with KCl, concentrations of 500~mM, 1~M, 2~M, and 3~M of 
were considered, corresponding to 52, 104, 208, or 311 pairs of K$^+$ and Cl$^-$ ions.

@@

The simulation parameters were taken from Ref.~\citenum{rog16}. 
Periodic boundary conditions were employed in all dimensions. The list of neighbors was updated every step. 
%\hl{I used nstlist=1 because it's used in the mdp in the article quoted. In DOI: 10.1021/jp5016627, nstlist=10 is used}
The Lennard-Jones interactions were cut off at 1~nm, whereas the smooth particle mesh Ewald algorithm 
\cite{essman95} was used for long-range electrostatics. Dispersion correction \cite{shirts07} 
was applied to both energy and pressure. Temperature was kept constant at 298~K using the 
Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}. The lipids and 
the solvent were coupled separately, and a time constant of 0.4~ps was employed. Pressure was 
maintained semi-isotropically at 1~bar using the Parrinello--Rahman barostat with a time constant of 1~ps 
\cite{parrinello81}. All non-water bonds were constrained with the LINCS algorithm 
\cite{hess97,hess07},
whereas the bonds in water molecules were constrained using SETTLE \cite{miyamoto92}.
The integration time step was set to 2~fs, and the simulations were performed either for 400~ns 
(only counterions), 300~ns (systems with KCl), or 600~ns (systems with CaCl$_2$). The last 
250, 200, or 300~ns of the simulations were included in the analyses, respectively.

\subsection{Lipid17}
{\it DOPS and POPS bilayers.} 
The structure of the bilayers with 128 DOPS or 128 POPS lipids were
obtained from CHARMM-GUI~\cite{lee16,jo18}. Bilayers are solvated with
4480 water molecules resulting in 31.1 water molecules per lipid. For
each system, 128 NaCl molecules are added to neutralize the system.
In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. SETTLE~\cite{miyamoto92} algorithm was employed 
to constrain the bonds in TIP3P water molecules. Simulations were ran using Amber18 simulation package \cite{amber18md}
and lipid parameters were obtained from Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with Joung--Cheatham
ion parameters~\cite{joung2008determination} and one set of simulations were run with ff99SB (\AA{}qvist)~\cite{aqvist90} ion parameters.
The data in Table~\ref{PSsystems} and Refs~\citenum{lipid17POPSjcions,lipid17POPSff99ions,lipid17DOPSjcions,lipid17DOPSff99ions} contain
the number of molecules in simulations, related simulation files and trajectories.

Each bilayer structure was first minimized for 2500 steps with steepest
descent and an additional 2500 step with the conjugate gradient
algorithms while restraining the solvent and ion molecules with a force
constant 500~kcal~mol$^{-1}$~\AA{}$^{-2}$. An additional minimization
step was used with the same parameters after removing the constraints on
the solvent and ion molecules. During the minimization procedure,
non-periodic boundary conditions were applied, no pressure scaling was
used and none of the bonds were restrained. A three step heating
procedure was applied: in the first heating step the system temperature
is increased from 0.0~K to 100.0~K using Langevin thermostat with
5.0~ps$^{-1}$ collision frequency while restraining the lipid positions
with harmonic springs of force constant
20.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at constant volume without periodic
boundary conditions. In the second heating step, the system temperature
was increased from 100.0~K to 200.0~K using Langevin thermostat with
5.0~ps$^{-1}$ collision frequency within 10000 steps while restraining
the lipid positions with harmonic springs of force constant
10.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at constant volume. In the third
heating step, the system temperature is increased from 200.0~K to
303.0~K within 10000 steps using Langevin thermostat with 5.0~ps$^{-1}$
collision frequency while restraining the lipid positions with harmonic
springs of force constant 10.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at 1~bar
constant pressure using Berendsen barostat with 3.0~ps relaxation time.
From the second heating step onward, periodic boundary conditions in all
dimensions were applied. The semi-isotropic pressure coupling along
\textit{xy}-plane was employed and the surface tension was set to
0.0~dyn explicitely. After the third heating step, each structure is
equilibriated for 100~ns at 1~bar and 303~K constant pressure using
Berendsen barostat and temperature using Langevin thermostat,
respectively. For the equilibriation step, semi-isotropic pressure
coupling along \textit{xy}-plane was employed and the surface tension
was set to 0.0~dyn explicitely. For all heating and equilibration steps,
the lengths of the hydrogen containing bonds were restrained using SHAKE
algorithm and a time step of 2~fs was used. In all steps with periodic
boundary conditions, the non-bonded interactions are calculated with
particle mesh Ewald~\cite{essman95} using 1.0~nm cutoff.

For the production runs, the temperature and pressure were set to 303~K
and 1~bar, using Langevin thermostat with 1.0~ps$^{-1}$ collision
frequency and Berendsen barostat with 1.0~ps relaxation time,
respectively. Periodic boundary conditions were applied in all
dimensions and the non-bonded interactions are calculated with particle
mesh Ewald~\cite{essman95} using 1.0~nm cutoff. The lengths of the
hydrogen containing bonds were constrained using SHAKE algorithm and a
time step of 2~fs was used. Trajectories are saved in every 10~ps
intervals. Two independent trajectories of 500~ns long was generated and
the last 200~ns portions were used for the analysis.

\noindent
{\it POPC:POPS (5:1) mixtures with additional potassium and sodium ions} 
The initial structure of the bilayers with 120 POPC and 24 POPS lipids were obtained from CHARMM-GUI~\cite{lee16,jo18}.
Bilayers were solvated with 5760 TIP3P water molecules~\cite{jorgensen1983comparison}. For the simulations of the lipids and ions,
the Amber Lipid 17 \cite{gould18} and ff99SB~\cite{aqvist90} force fields were used, respectively.
Simulations were ran using Amber18 simulation package \cite{amber18md} 
The simulation steps are the same as the the simulations of
DOPS and POPS bilayers described above, with the only difference that the total simulation time was 300~ns, of which last 200~ns
was used for the analysis. The number of ions in each specific system is given in Table~\ref{mixedIONsystemsMONOVALENT}.
The trajectory files can be found in Refs~\citenum{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa}.

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration with default Amber parameters} 
%The data in Table \ref{mixedIONsystems} and Refs \citenum{lipid17_cacl_series,POPCpopsLIPID17withCaCl} 
%We have performed two sets of simulations of POPC:POPS (5:1) mixtures using Amber Lipid17 force field \cite{gould18}
%with additional calcium chloride concentrations, 
%which employed different simulation engines and parameters for ions.
Simulations were performed using Amber18 simulation package \cite{amber18md} with ff99SB~\cite{aqvist90} ion parameters and Amber Lipid17~\cite{gould18} force field. 
Number of ions and water molecules for this set of simulations are shown in Table~\ref{mixedIONsystems}.
The simulation details, minimization, heating, and equilibration runs are the same as described above for the other POPC:POPS (5:1) simulations with Amber Lipid17 forcefield. 
For each ion concentration, 100~ns equilibration runs were followed by 300~ns production runs with two independent trajectories, of which last 200~ns were used for the analysis.
The trajectory files can be found from Ref~\citenum{POPCpopsLIPID17withCaCl}.

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration with Dang ion parameters} 
The Lipid17 parameters for POPS were obtained from AmberTools18 \cite{amber18md} 
and converted to Gromacs format using acpype \cite{silva12}.  
The previously generated \cite{botan15} Lipid14 parameters for POPC in Gromacs format were downloaded from Ref.~\citenum{lipid14files}. 
Dang parameters for ions \cite{smith94, dang06} were used.
% Another set of simulations with Lipid17 force field for lipids \cite{gould18}
Initial structures were generated using CHARMM-GUI~\cite{lee16,jo18} 
and simulations were performed using the GROMACS 2018 simulation package~\cite{abraham2015gromacs}
with the time step of 2~fs. 
The non-bonded interactions were calculated directly within 1.0~nm cutoff using the Verlet scheme \cite{Pall13}, 
and the long-range electrostatic forces were calculated using particle mesh Ewald~\cite{essman95}. 
The bond lengths of hydrogen atoms were constrained using LINCS algorithm \cite{hess97}.
Temperature was coupled to the velocity rescaling thermostat \cite{bussi07} at 298~K with a coupling constant of 1~ps.
Pressure was coupled to the Parrinello--Rahman barostat \cite{parrinello81} at 1~bar with a coupling constant of 10~ps. 
The compositions of individual simulations are shown in Table~\ref{mixedIONsystems}. 
The systems were equilibrated 50~ns prior to production simulations used for analysis,
which are available in Ref.~\citenum{lipid17_cacl_series} including the simulation settings. 
Topologies in a format for the GROMACS package are available in the repository \citenum{ECCpops}.

\subsection{Slipids}
%
%\noindent
%{\it POPS}
%\todo{POPS \cite{slipidsPOPS298K} to be written by Piggot}%%
%
%\noindent
%{\it DOPS with 128 lipids.}
%\todo{DOPS with 128 lipids \cite{slipidsDOPS303K} to be written by Piggot}
%
\noindent
{\it POPS and DOPS with 128 lipids}
Starting structures for the Slipids POPS and DOPS (with 128 lipids) simulations were the same as constructed
for the CHARMM36 simulations. Force field parameters were taken from the published Slipids PS model~\cite{jambeck13},
however simulation cut-offs used herein were different to these previously published simulations~\cite{jambeck2012another}.
% [REF: https://pubs.acs.org/doi/10.1021/ct300777p].
This was due to the use of the older GROMACS group-based cut-off scheme in these previous simulations.
In the simulations reported in this work, the Verlet cut-off scheme was used. Van der Waals interactions were
truncated at 1.0 nm with LJ-PME~\cite{wennberg15}
% [REF: https://pubs.acs.org/doi/10.1021/acs.jctc.5b00726]
applied to account for the long-range interactions.
Coulombic interactions were truncated at 1.0 nm with PME used for the long-range interactions.
Full testing and validation of these settings with the Slipids PS force field will be provided in an additional forthcoming work.
All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations.

@@

\noindent
{\it POPC:POPS (5:1) mixtures without additional ions}
% The data in Table \ref{mixedIONsystems} and Refs. \citenum{POPCpopsGROMOSCKPwithNa,POPCpopsGROMOSCKPMwithNa}
% \todo{T. Piggot, please write simulation details} \\
Starting structures for GROMOS-CKP and GROMOS-CKPM simulations were taken from those produced for the equivalent CHARMM36 simulations (see above).
A script was used to re-order the atoms and to remove any extraneous hydrogen atoms. Simulation setting were identical to those as described above
for the GROMOS-CKP(M) POPS bilayers.

>>>>

Simulations of these systems were performed for 500~ns using GROMACS, version 5.0.6 \cite{abraham2015gromacs}.
A 2~fs timestep was applied during the simulations, and the LINCS algorithm was used to constrain all bonds to hydrogen atoms \cite{hess97,hess07}.
For each of the DOPS and POPS systems, two simulations were performed using different randomly assigned starting velocities.
The simulations were maintained at temperatures of 303~K and 298~K for DOPS and POPS respectively using the Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}
with a coupling constant of 1~ps. A pressure of 1~bar was maintained using the Parrinello--Rahman \cite{parrinello81} method with a coupling constant of 5~ps.
Pressure coupling was applied in a semi-isotropic manner with the $x$ and $y$ dimensions, in the plane of the bilayer, fluctuating independently of the $z$ dimension.
Standard CHARMM36 force field methods were used for the simulation cut-offs: van der Waals interactions were truncated at 1.2~nm with the interactions
switched off between 1.0~nm and 1.2~nm; Coulombic interactions were truncated at 1.2~nm with long-range interactions treated using PME~\cite{darden93,essman95}. The final 100~ns of the simulations was used for analysis.

\noindent
{\it POPC:POPS mixtures without additional ions.}
Inputs for simulations of POPC:POPS (5:1) mixtures containing a total of 132 lipids (110 POPC and 22 POPS) were configured using the CHARMM-GUI~\cite{lee16,jo18}. Two identical systems were constructed, apart from the counter ion; one system was neutralised through the addition of 22 Na$^+$ ions while the other system
contained 22 K$^+$ ions. As in the pure POPS CHARMM36 simulations, 35 water per lipid were added. All other system and simulation parameters
and settings were identical to those used in the CHARMM36 POPS simulations.

@@

Simulations of POPC:POPS (5:1) ans POPC:POPS (1:1) mixtures containing total 300 lipids (250:50 and 150:150, respectively)
with neutralizing potassium countertions 
%ran using CHARMM36 force field.
%The data are available from Refs. \citenum{POPC5POPS1noCaClCHARMM} and \citenum{POPC1POPS1noCaClCHARMM}.
%POPC:POPS (both 1:1 and 5:1) mixtures
were prepared
%with 300 lipids in total
%with the  ions
using the CHARMM-GUI \cite{lee16,jo18}. The systems were simulated using Gromacs 5 \cite{abraham2015gromacs}
and CHARMM36 force field with the simulation parameters given by the CHARMM-GUI at 298~K. For further details see table \ref{mixedIONsystemsMONOVALENT}.

\noindent
{\it POPC:POPS (5:1) mixtures with additional monovalent ions.}
POPC:POPS (110:22) mixtures containing total 132 lipids with the additional
potassium or sodium ions corresponding concentrations of $\sim$450~mM and 890~mM
were generated with the CHARMM-GUI \cite{lee16,jo18}. Systems were simulated using
Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation
parameters given by the CHARMM-GUI at 298~K. For further details, simulation files
and data, see table \ref{mixedIONsystemsMONOVALENT} and
Refs. \citenum{charmm36pops+83popcT298Kwith450mMK,charmm36pops+83popcT298Kwith890mMK,charmm36pops+83popcT298Kwith440mMNa,charmm36pops+83popcT298Kwith890mMNa}.

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium using NBfix1 parameters.} 
POPC:POPS (5:1) mixtures containing total 300 (250:50) lipids
%were simulated with the additional calcium concentrations.
%The data is available from Refs. \citenum{POPC5POPS1withCaClCHARMM} and \citenum{POPC5POPS1with1MCaClCHARMM}.
and 0.26~M and 1~M CaCl$_2$ were prepared using the CHARMM-GUI in January 2018 \cite{lee16,jo18}.
The ion parameters generated by CHARMM-GUI at the time are labeled here as NBfix1~\cite{kim16}.
Systems were simulated using Gromacs 5 \cite{abraham2015gromacs} and CHARMM36 force field with the simulation parameters given
by the CHARMM-GUI at 298~K.

@@

\noindent
{\it POPS bilayers with Na$^+$ counterions.}
%The data in Table \ref{PSsystems} and Refs \citenum{macrogPOPS298Kcorrect,macrogPOPS298KwithK}
%\todo{T. Piggot and M. Javanainen, please write simulation details} \\
%\hl{Piggot: please write about system with Na$^+$ counterions}.
Force field parameters for POPS simulations with the MacRog OPLS-AA force field~\cite{maciejewski14,kulig15b,rog16}
were constructed manually using the PS parameters provided in ref.~\citenum{rog16} as a guide. The original parameters were
not used due to an incorrect arrangement of the {\it sn}-1 and {\it sn}-2 tails (i.e. the lipid was OPPS and not POPS).
The corrected parameters are available from Ref.~\citenum{macrogPOPS298KwithK}.
The starting structure was the same as that used in the CHARMM36 POPS simulations (with Na$^+$ ions) and was converted into
the correct format using a script to adjust the atom order. Water was treated using the TIP3P model~\cite{jorgensen83}.
Simulation parameters were chosen to closely mimic those used in the original force field publications~\cite{maciejewski14,kulig15b,rog16}:
van der Waals interactions were truncated at 1.0~nm with a dispersion correction applied to the energy and pressure;
Long-range Coulombic interactions were treated with PME beyond a system-optimized cut-off.
All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations,
including use of the Verlet cut-off scheme~\cite{Pall13}, apart from the use of LINCS to constrain all bonds during these simulations.
Simulations with the group cut-off scheme gave slightly lower area per molecule ($\sim$~50\AA$^2$) than
the simulations with Verlet cut-off ($\sim$~52\AA$^2$).

@@

\noindent
{\it POPC:POPS (5:1) mixtures with additional potassium ions.} 
%The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsMACROG,POPCpopsMACROGwithK}
The bilayers containing a total of 120 POPC and 24 POPS molecules distributed evenly among the 
two leaflets were set up using CHARMM-GUI \cite{lee16,jo18}.
The bilayers were solvated by 5760 water molecules, corresponding to 40 water molecules for lipid.
Additionally, 24 potassium ions were added to neutralize the charge of the POPS head groups.
These bilayers were simulated using Gromacs 2016.3~\cite{abraham2015gromacs} in the presence of counterions only, as well as together with different
concentrations of CaCl$_2$ or KCl. For the former, concentrations of 100~mM, 300~mM, 1~M, and 3~M 
were considered, which corresponded to the amounts of 10/20, 31/62, 104/208, and 311/622 
of Ca$^{2+}$/Cl$^-$ ions. For membranes with KCl, concentrations of 500~mM, 1~M, 2~M, and 3~M of 
were considered, corresponding to 52, 104, 208, or 311 pairs of K$^+$ and Cl$^-$ ions.

@@

The simulation parameters were taken from Ref.~\citenum{rog16}. 
Periodic boundary conditions were employed in all dimensions. The neighbor list was updated every step. 
%\hl{I used nstlist=1 because it's used in the mdp in the article quoted. In DOI: 10.1021/jp5016627, nstlist=10 is used}
The Lennard-Jones interactions were cut off at 1~nm, whereas the smooth particle mesh Ewald algorithm 
\cite{essman95} was used for long-range electrostatics. Dispersion correction \cite{shirts07} 
was applied to both energy and pressure. Temperature was kept constant at 298~K using the 
Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}. The lipids and 
the solvent were coupled separately, and a time constant of 0.4~ps was employed. Pressure was 
maintained semi-isotropically at 1~bar using the Parrinello--Rahman barostat with a time constant of 1~ps 
\cite{parrinello81}. All non-water bonds were constrained with the LINCS algorithm 
\cite{hess97,hess07},
whereas the bonds in water molecules were constrained using SETTLE \cite{miyamoto92}.
The integration time step was set to 2~fs, and the simulations were performed either for 400~ns 
(only counterions), 300~ns (systems with KCl), or 600~ns (systems with CaCl$_2$). The last 
250, 200, or 300~ns of the simulations were included in the analyses, respectively.

\subsection{Lipid17}
{\it DOPS and POPS bilayers.} 
The structure of the bilayers with 128 DOPS or 128 POPS lipids were
obtained from CHARMM-GUI~\cite{lee16,jo18}. Bilayers are solvated with
4480 water molecules resulting in 31.1 waters per lipid. For
each system, 128 NaCl molecules were added to neutralize the system.
In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. SETTLE~\cite{miyamoto92} algorithm was employed 
to constrain the bonds in TIP3P water molecules. Simulations were run using Amber18 simulation package \cite{amber18md}
and lipid parameters were obtained from Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with Joung--Cheatham
ion parameters~\cite{joung2008determination} and another set of simulations were run with ff99SB (\AA{}qvist)~\cite{aqvist90} ion parameters.
The data in Table~\ref{PSsystems} and Refs~\citenum{lipid17POPSjcions,lipid17POPSff99ions,lipid17DOPSjcions,lipid17DOPSff99ions} contain
the number of molecules in simulations, related simulation files and trajectories.

Each bilayer structure was first minimized for 2500 steps using the steepest
descent algorithm and then an additional 2500 steps with the conjugate gradient
algorithms while restraining the solvent and ions with a force
constant 500~kcal~mol$^{-1}$~\AA{}$^{-2}$. An additional minimization
step was used with the same parameters after removing the constraints on
the solvent and ion molecules. During the minimization procedure,
non-periodic boundary conditions were applied, no pressure scaling was
used and none of the bonds were restrained. A three step heating
procedure was applied: in the first heating step the system temperature
is increased from 0.0~K to 100.0~K using Langevin thermostat with
5.0~ps$^{-1}$ collision frequency while restraining the lipid positions
with harmonic springs of force constant
20.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at constant volume without periodic
boundary conditions. In the second heating step, the system temperature
was increased from 100.0~K to 200.0~K using Langevin thermostat with
5.0~ps$^{-1}$ collision frequency within 10000 steps while restraining
the lipid positions with harmonic springs of force constant
10.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at constant volume. In the third
heating step, the system temperature is increased from 200.0~K to
303.0~K within 10000 steps using Langevin thermostat with 5.0~ps$^{-1}$
collision frequency while restraining the lipid positions with harmonic
springs of force constant 10.0~kcal~mol$^{-1}$~\AA{}$^{-1}$ at 1~bar
constant pressure using Berendsen barostat with 3.0~ps relaxation time.
From the second heating step onward, periodic boundary conditions in all
dimensions were applied. The semi-isotropic pressure coupling along
\textit{xy}-plane was employed and the surface tension was set to
0.0~dyn explicitely. After the third heating step, each structure is
equilibriated for 100~ns at 1~bar and 303~K constant pressure using
Berendsen barostat and temperature using Langevin thermostat,
respectively. For the equilibriation step, semi-isotropic pressure
coupling along \textit{xy}-plane was employed and the surface tension
was set to 0.0~dyn explicitely. For all heating and equilibration steps,
the lengths of the hydrogen containing bonds were restrained using SHAKE
algorithm and a time step of 2~fs was used. In all steps with periodic
boundary conditions, the non-bonded interactions were calculated with
particle mesh Ewald~\cite{essman95} using 1.0~nm cutoff.

For the production runs, the temperature and pressure were set to 303~K
and 1~bar, using Langevin thermostat with 1.0~ps$^{-1}$ collision
frequency and Berendsen barostat with 1.0~ps relaxation time,
respectively. Periodic boundary conditions were applied in all
dimensions and the non-bonded interactions are calculated with particle
mesh Ewald~\cite{essman95} using 1.0~nm cutoff. The lengths of the
hydrogen containing bonds were constrained using SHAKE algorithm and a
time step of 2~fs was used. Trajectories are saved in every 10~ps
intervals. Two independent, 500~ns long trajectories were generated, of which
the last 200~ns was used for the analysis.

\noindent
{\it POPC:POPS (5:1) mixtures with additional potassium and sodium ions.} 
The initial structure of the bilayers with 120 POPC and 24 POPS lipids were obtained from CHARMM-GUI~\cite{lee16,jo18}.
Bilayers were solvated with 5760 TIP3P water molecules~\cite{jorgensen1983comparison}. For the simulations of the lipids and ions,
the Amber Lipid17 \cite{gould18} and ff99SB~\cite{aqvist90} force fields were used, respectively.
Simulations were ran using Amber18 simulation package \cite{amber18md} 
The simulation steps are the same as the the simulations of
DOPS and POPS bilayers described above, with the only difference that the total simulation time was 300~ns, of which last 200~ns
was used for the analysis. The number of ions in each specific system is given in Table~\ref{mixedIONsystemsMONOVALENT}.
The trajectory files can be found in Refs~\citenum{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa}.

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration using default Amber parameters.} 
%The data in Table \ref{mixedIONsystems} and Refs \citenum{lipid17_cacl_series,POPCpopsLIPID17withCaCl} 
%We have performed two sets of simulations of POPC:POPS (5:1) mixtures using Amber Lipid17 force field \cite{gould18}
%with additional calcium chloride concentrations, 
%which employed different simulation engines and parameters for ions.
Simulations were performed using Amber18 simulation package \cite{amber18md} with ff99SB~\cite{aqvist90} ion parameters and Amber Lipid17~\cite{gould18} force field. 
Number of ions and water molecules for this set of simulations are shown in Table~\ref{mixedIONsystems}.
The simulation details, minimization, heating, and equilibration runs were the same as described above for the other POPC:POPS (5:1) simulations with Amber Lipid17 forcefield. 
For each ion concentration, 100~ns equilibration runs were followed by 300~ns production runs with two independent trajectories, of which last 200~ns were used for the analysis.
The trajectory files can be found from Ref~\citenum{POPCpopsLIPID17withCaCl}.

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium chloride concentration using Dang ion parameters.} 
The Lipid17 parameters for POPS were obtained from AmberTools18 \cite{amber18md} 
and converted to Gromacs format using acpype \cite{silva12}.  
The previously generated \cite{botan15} Lipid14 parameters for POPC in Gromacs format were downloaded from Ref.~\citenum{lipid14files}. 
Dang parameters for ions \cite{smith94, dang06} were used.
% Another set of simulations with Lipid17 force field for lipids \cite{gould18}
Initial structures were generated using CHARMM-GUI~\cite{lee16,jo18} 
and simulations were performed using the GROMACS 2018 simulation package~\cite{abraham2015gromacs}
with the time step of 2~fs. 
The non-bonded interactions were calculated directly within 1.0~nm cutoff using the Verlet scheme \cite{Pall13}, 
and the long-range electrostatic forces were calculated using particle mesh Ewald~\cite{essman95}. 
The bond lengths of hydrogen atoms were constrained using LINCS algorithm \cite{hess97}.
Temperature was coupled to the velocity rescaling thermostat \cite{bussi07} at 298~K with a coupling constant of 1~ps.
Pressure was coupled to the Parrinello--Rahman barostat \cite{parrinello81} at 1~bar with a coupling constant of 10~ps. 
The compositions of individual simulations are shown in Table~\ref{mixedIONsystems}. 
The systems were equilibrated 50~ns prior to production simulations used for analysis,
which are available in Ref.~\citenum{lipid17_cacl_series} including the simulation settings. 
Topologies in a format for the GROMACS package are available in the repository \citenum{ECCpops}.

\subsection{Slipids}
%
%\noindent
%{\it POPS}
%\todo{POPS \cite{slipidsPOPS298K} to be written by Piggot}%%
%
%\noindent
%{\it DOPS with 128 lipids.}
%\todo{DOPS with 128 lipids \cite{slipidsDOPS303K} to be written by Piggot}
%
\noindent
{\it POPS and DOPS with 128 lipids.}
Starting structures for the Slipids POPS and DOPS (with 128 lipids) simulations were the same as constructed
for the CHARMM36 simulations. Force field parameters were taken from the published Slipids PS model~\cite{jambeck13},
however simulation cut-offs used herein were different to these previously published simulations~\cite{jambeck2012another}.
% [REF: https://pubs.acs.org/doi/10.1021/ct300777p].
This was due to the use of the older GROMACS group-based cut-off scheme in these previous simulations.
In the simulations reported in this work, the Verlet cut-off scheme was used. Van der Waals interactions were
truncated at 1.0 nm with LJ-PME~\cite{wennberg15}
% [REF: https://pubs.acs.org/doi/10.1021/acs.jctc.5b00726]
applied to account for the long-range interactions.
Coulombic interactions were truncated at 1.0 nm with PME used for the long-range interactions.
Full testing and validation of these settings with the Slipids PS force field will be provided in an additional forthcoming work.
All other simulation settings were identical to those described above for the CHARMM36 DOPS and POPS simulations.

@@

\noindent
{\it POPC:POPS (5:1) mixtures without additional ions.}
% The data in Table \ref{mixedIONsystems} and Refs. \citenum{POPCpopsGROMOSCKPwithNa,POPCpopsGROMOSCKPMwithNa}
% \todo{T. Piggot, please write simulation details} \\
Starting structures for GROMOS-CKP and GROMOS-CKPM simulations were taken from those produced for the equivalent CHARMM36 simulations (see above).
A script was used to re-order the atoms and to remove any extraneous hydrogen atoms. Simulation setting were identical to those as described above
for the GROMOS-CKP(M) POPS bilayers.
